---
title: "Association between Audio Features, Popularity, and Listeners’ Preferences"
author: "Avianna Bui"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Part 1. Data Preprocessing

## Load Packages

```{r, message=F, warning = FALSE}
library(dplyr)
library(readr)
library(ggplot2)
library(rpart.plot) 
library(tidymodels)
library(lubridate)
library(cluster)
library(purrr)
library(ISLR2)
library(plotmo)
library(viridis) 
library(vip)
library(corrplot)
library(ggcorrplot)
library(devtools)
library(factoextra)
library(pander)
library(ggpubr)
tidymodels_prefer()
```

## Load Data

```{r, message = F}
billboard <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-09-14/billboard.csv')

audio_features <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-09-14/audio_features.csv')
```

## Data Cleaning

```{r, message = F, warning = F}
songs <- billboard %>%
  left_join(audio_features) %>%
  mutate(year = year(as.Date(week_id, "%m/%d/%Y"))) %>%
  select(year, song, performer, spotify_track_duration_ms, danceability, energy, key, loudness, mode, speechiness, acousticness, instrumentalness, liveness, valence, tempo, time_signature, spotify_track_popularity) %>%
  drop_na() %>%
  distinct(year, song, performer, .keep_all = TRUE) %>%
  mutate(across(where(is.character), as.factor)) %>%
  arrange(year)
```

# Part 2. Supervised Learning: Predict Spotify Track Popularity from Audio Features

## Data Cleaning

```{r}
songs_sl <- songs %>%
  select(-c(performer, year))
```

## Correlation Matrix 

```{r}
ggcorrplot(cor(songs_sl %>% select(-c(song, spotify_track_popularity))),
           lab = TRUE,
           ggtheme = ggplot2::theme_bw,
           colors = c("#ffff9f", "#48aaad", "#0a1172"),
           lab_size = 2, 
           legend.title = "Correlation") +
  theme(axis.text.x=element_text(size=8),
        axis.text.y = element_text(size = 8),
        legend.text = element_text(size = 8),
        legend.title = element_text(size = 10),
        plot.background = element_rect(fill = "black", color = "black"), panel.background = element_rect(color = "black", fill = "black"), text = element_text(color = "white"), axis.text=element_text(color="white"), axis.line =element_line(color = "white"), legend.background = element_rect(fill = "black", color = "black"))
```

## LASSO Modeling

```{r}
set.seed(202)

songs_cv <- vfold_cv(songs_sl, v = 10) 

data_rec <- recipe(spotify_track_popularity ~ ., data = songs_sl) %>%
    update_role(song, new_role = "ID") %>%
    step_nzv(all_predictors()) %>%
    step_normalize(all_numeric_predictors()) 

lm_lasso_spec_tune <- 
    linear_reg() %>%
    set_args(mixture = 1, penalty = tune()) %>% 
    set_engine(engine = "glmnet") %>% 
    set_mode("regression") 

lasso_wf_tune <- workflow() %>% 
    add_recipe(data_rec) %>% 
    add_model(lm_lasso_spec_tune) 

penalty_grid <- grid_regular(
    penalty(range = c(-1, 3)), 
    levels = 40)

tune_output <- tune_grid( 
    lasso_wf_tune,
    resamples = songs_cv,
    metrics = metric_set(mae),
    grid = penalty_grid)

tune_output %>% collect_metrics(summarize = TRUE)
```

```{r}
# select the "best" value of lambda
best_se_penalty <- select_by_one_std_err(tune_output, metric = "mae", desc(penalty)) 
best_se_penalty
```

```{r}
# select the best model
final_wf_se <- finalize_workflow(lasso_wf_tune, best_se_penalty)

final_fit_se <- fit(final_wf_se, data = songs_sl)

tidy(final_fit_se)
```

```{r}
# visualize coefficient path graph
glmnet_output <- final_fit_se %>% extract_fit_parsnip() %>% pluck("fit")

plot_glmnet(glmnet_output, label = TRUE, col = viridis(13)) 
```

```{r}
# arrange variables based on their LASSO importance
bool_predictor_exclude <- glmnet_output$beta==0

var_imp <- sapply(seq_len(nrow(bool_predictor_exclude)), function(row) {
    this_coeff_path <- bool_predictor_exclude[row,]
    if(sum(this_coeff_path) == ncol(bool_predictor_exclude)){ return(0)}else{
    return(ncol(bool_predictor_exclude) - which.min(this_coeff_path) + 1)}
})

var_imp_data <- tibble(
    var_name = rownames(bool_predictor_exclude),
    var_imp = var_imp
)
var_imp_data %>% arrange(desc(var_imp))
```

## Generalized Additive Model (GAM)

```{r, message = F}
# residual plot: model selection

lasso_mod_out <- final_fit_se %>%
    predict(new_data = songs_sl) %>%
    bind_cols(songs_sl) %>%
    mutate(resid = spotify_track_popularity - .pred)

residplot1 <- lasso_mod_out %>% 
    ggplot(aes(x = .pred, y = resid)) + 
    geom_point(alpha = 0.2, color = "#f1abb9") +
    geom_smooth(color = "#7e6551", se = FALSE) + 
    geom_hline(yintercept = 0, color = "#f4f1de") + 
    theme_classic() +
    theme(plot.background = element_rect(fill = "black", color = "black"), panel.background = element_rect(color = "black", fill = "black"), text = element_text(color = "white"), axis.text=element_text(color="white"), axis.line =element_line(color = "white"))

residplot2 <- lasso_mod_out %>% 
    ggplot(aes(x = loudness, y = resid)) + 
    geom_point(alpha = 0.2, color = "#f1abb9") +
    geom_smooth(color = "#7e6551", se = FALSE) + 
    geom_hline(yintercept = 0, color = "#f4f1de") + 
    theme_classic() +
    theme(plot.background = element_rect(fill = "black", color = "black"), panel.background = element_rect(color = "black", fill = "black"), text = element_text(color = "white"), axis.text=element_text(color="white"), axis.line =element_line(color = "white"))

# png(filename="residplots.png", width=1600, height=600)
ggarrange(residplot1, residplot2)
```

```{r, message = F, warning = F}
gam_spec <- 
    gen_additive_mod() %>%
    set_args(select_features = TRUE, adjust_deg_free = tune()) %>%
    set_engine(engine = "mgcv") %>%
    set_mode("regression")

gam_wf <- workflow() %>% 
    add_variables(outcomes = spotify_track_popularity, predictors = c(spotify_track_duration_ms, danceability, energy, loudness, mode, speechiness, acousticness, instrumentalness, liveness, valence, tempo, time_signature)) %>% 
    add_model(gam_spec, formula = spotify_track_popularity ~ mode + s(spotify_track_duration_ms, k=2) + s(danceability, k=2) + s(energy, k=2) + s(loudness, k=2) + s(speechiness, k=2) + s(acousticness, k=2) + s(instrumentalness, k=2) + s(liveness, k=2) + s(valence, k=2) + s(tempo, k=2) + s(time_signature, k=2))

tuning_param_grid <- grid_regular(
    adjust_deg_free(range = c(0.25, 4)),
    levels = 8
)

tune_output <- tune_grid( 
    gam_wf,
    resamples = songs_cv,
    metrics = metric_set(mae),
    grid = tuning_param_grid
)
```

```{r, warning=F}
tune_output %>% collect_metrics()
tune_output %>% show_best()

autoplot(tune_output) + theme_classic()

best_by_1se_param <- tune_output %>% 
    select_by_one_std_err(metric = "mae", desc(adjust_deg_free))

best_by_1se_param 

gam_mod_best1se <- finalize_workflow(gam_wf, best_by_1se_param) %>%
    fit(data = songs_sl)
```

## Random Forest

```{r}
set.seed(202)
rand_forest <- songs_sl %>%
  select(-c(song))
  
rf_spec <- rand_forest() %>%
    set_engine(engine = "ranger") %>% 
    set_args(
        mtry = NULL, 
        trees = 500, 
        min_n = 2,
        importance = "impurity"
    ) %>%
    set_mode("regression") 

data_rec <- recipe(spotify_track_popularity ~ ., data = rand_forest)

data_wf <- workflow() %>%
    add_model(rf_spec) %>%
    add_recipe(data_rec)

rf_fit <- fit(data_wf, data = rand_forest)


rf_fit
```

```{r}
# variable importance
rf_fit %>% 
    extract_fit_engine() %>% 
    vip(num_features = 13, aesthetics = list(color = "pink", fill = "pink")) + 
    theme_classic() +
    theme(plot.background = element_rect(fill = "black", color = "black"), panel.background = element_rect(color = "black", fill = "black"), text = element_text(color = "white"), axis.text=element_text(color="white"), axis.line =element_line(color = "white"))
```

## Evaluate Model Performance and Variable Importance

```{r}
songs %>%
  mutate(year = factor(year)) %>%
  ggplot(aes(x = year, y = spotify_track_popularity)) +
  geom_boxplot(fill = "#ffcfe3", color = "#ff9bc5") +
  geom_hline(yintercept = mean(songs$spotify_track_popularity), color = "white", linetype = "dotdash") +
  theme_classic() +
  theme(plot.background = element_rect(fill = "black", color = "black"), panel.background = element_rect(color = "black", fill = "black"), legend.background = element_rect(color = "black", fill = "black"), text = element_text(color = "white"), axis.text=element_text(color="white"), axis.line =element_line(color = "white"), axis.text.x =  element_text(angle = 90, size = 8))
```

```{r, message = F}
scat1 <- songs_sl %>% 
    ggplot(aes(x = loudness, y = spotify_track_popularity)) + 
    geom_point(alpha = 0.2, color = "#f1abb9") +
    geom_smooth(color = "white", se = FALSE) + 
    theme_classic() +
    theme(plot.background = element_rect(fill = "black", color = "black"), panel.background = element_rect(color = "black", fill = "black"), text = element_text(color = "white"), axis.text=element_text(color="white"), axis.line =element_line(color = "white"))

scat2 <- songs_sl %>% 
    ggplot(aes(x = valence, y = spotify_track_popularity)) + 
    geom_point(alpha = 0.2, color = "#f1abb9") +
    geom_smooth(color = "white", se = FALSE) + 
    theme_classic() +
    theme(plot.background = element_rect(fill = "black", color = "black"), panel.background = element_rect(color = "black", fill = "black"), text = element_text(color = "white"), axis.text=element_text(color="white"), axis.line =element_line(color = "white"))

scat3 <- songs_sl %>% 
    ggplot(aes(x = acousticness, y = spotify_track_popularity)) + 
    geom_point(alpha = 0.2, color = "#f1abb9") +
    geom_smooth(color = "white", se = FALSE) + 
    theme_classic() +
    theme(plot.background = element_rect(fill = "black", color = "black"), panel.background = element_rect(color = "black", fill = "black"), text = element_text(color = "white"), axis.text=element_text(color="white"), axis.line =element_line(color = "white"))

scat4 <- songs_sl %>% 
    ggplot(aes(x = speechiness, y = spotify_track_popularity)) + 
    geom_point(alpha = 0.2, color = "#f1abb9") +
    geom_smooth(color = "white", se = FALSE) + 
    theme_classic() +
    theme(plot.background = element_rect(fill = "black", color = "black"), panel.background = element_rect(color = "black", fill = "black"), text = element_text(color = "white"), axis.text=element_text(color="white"), axis.line =element_line(color = "white"))

# png(filename="scatterplots.png", width=1600, height=1000)
ggarrange(scat1, scat2, scat3, scat4, ncol = 2, nrow = 2)
```

# Part 3. Unsupervised Learning: Identify Song Clusters across Decades among Important Audio Features

## Data Cleaning

```{r}
songs_2000s <- songs %>%
  filter(year >= 2000 & year <= 2009) %>%
  select(valence, acousticness, loudness)

songs_2010s <- songs %>%
  filter(year >= 2010 & year <= 2019) %>%
  select(valence, acousticness, loudness)
```

## Picking $k$
```{r}
tot_wc_ss_2000 <- rep(0, 15)

for (k in 1:15) {
    kclust_2000 <- kmeans(scale(songs_2000s), centers = k)

    tot_wc_ss_2000[k] <- kclust_2000$tot.withinss
}

tot_wc_ss_2010 <- rep(0, 15)


for (k in 1:15) {
    # Perform clustering
    kclust_2010 <- kmeans(scale(songs_2010s), centers = k)

    # Store the total within-cluster sum of squares
    tot_wc_ss_2010[k] <- kclust_2010$tot.withinss
}
```


```{r}
# png(filename="k_plot.png", width=1600, height=600)

par(bg = "black", mfrow = c(1, 2))
plot(1:15, tot_wc_ss_2000, type = "b", main = "2000s", xlab = "Number of clusters", ylab = "Total within-cluster sum of squares", col = "pink",           
    col.main = "white",    
    col.lab = "white",    
    col.axis = "white",   
    fg = "white")

plot(1:15, tot_wc_ss_2010, type = "b", main = "2010s", xlab = "Number of clusters", ylab = "Total within-cluster sum of squares", col = "pink",           
    col.main = "white",   
    col.lab = "white",    
    col.axis = "white",   
    fg = "white") 
```

## K-Means Clustering

```{r}
set.seed(202)
kclust_2000_scale <- kmeans(scale(songs_2000s), centers = 6)
kclust_2000_scale

kclust_2010_scale <- kmeans(scale(songs_2010s), centers = 6)
kclust_2010_scale

songs_2000s <- songs_2000s %>%
    mutate(
        kclust_2000_scale = factor(kclust_2000_scale$cluster)
    )

songs_2010s <- songs_2010s %>%
    mutate(
        kclust_2010_scale = factor(kclust_2010_scale$cluster)
    )
```
## Cluster Visualization

```{r, message = F}
# loudness vs acousticness
ggplot(songs_2000s, aes(y = acousticness, x = loudness, color = kclust_2000_scale)) +
    geom_point() +
    geom_smooth(se = FALSE) +
    scale_color_brewer(palette = "Set3") +
    labs(title = "2000s") +
    theme_classic() +
    theme(plot.background = element_rect(fill = "black", color = "black"), panel.background = element_rect(color = "black", fill = "black"), text = element_text(color = "white"), axis.text=element_text(color="white"), axis.line =element_line(color = "white"), legend.position = "none", plot.title = element_text(face = "bold", hjust = 0.5))


ggplot(songs_2010s, aes(y = acousticness, x = loudness, color = kclust_2010_scale)) +
    geom_point() +
    geom_smooth(se = FALSE) +
    scale_color_brewer(palette = "Set3") +
    labs(title = "2010s") +
    theme_classic() +
    theme(plot.background = element_rect(fill = "black", color = "black"), panel.background = element_rect(color = "black", fill = "black"), text = element_text(color = "white"), axis.text=element_text(color="white"), axis.line =element_line(color = "white"), legend.position = "none", plot.title = element_text(face = "bold", hjust = 0.5))
```

```{r, message=F}
# loudness vs valence
ggplot(songs_2000s, aes(y = loudness, x = valence, color = kclust_2000_scale)) +
    geom_point() +
    geom_smooth(se = FALSE) +
    scale_color_brewer(palette = "Set3") +
    labs(title = "2000s") +
    theme_classic() +
    theme(plot.background = element_rect(fill = "black", color = "black"), panel.background = element_rect(color = "black", fill = "black"), text = element_text(color = "white"), axis.text=element_text(color="white"), axis.line =element_line(color = "white"), legend.position = "none", plot.title = element_text(face = "bold", hjust = 0.5))

ggplot(songs_2010s, aes(y = loudness, x = valence, color = kclust_2010_scale)) +
    geom_point() +
    geom_smooth(se = FALSE) +
    scale_color_brewer(palette = "Set3") +
    labs(title = "2010s") +
    theme_classic() +
    theme(plot.background = element_rect(fill = "black", color = "black"), panel.background = element_rect(color = "black", fill = "black"), text = element_text(color = "white"), axis.text=element_text(color="white"), axis.line =element_line(color = "white"), legend.position = "none", plot.title = element_text(face = "bold", hjust = 0.5))
```

```{r, message = F}
# valence vs acousticness
ggplot(songs_2000s, aes(y = valence, x = acousticness, color = kclust_2000_scale)) +
    geom_point() +
    geom_smooth(se = FALSE) +
    scale_color_brewer(palette = "Set3") +
    labs(title = "2000s") +
    theme_classic() +
    theme(plot.background = element_rect(fill = "black", color = "black"), panel.background = element_rect(color = "black", fill = "black"), text = element_text(color = "white"), axis.text=element_text(color="white"), axis.line =element_line(color = "white"), legend.position = "none", plot.title = element_text(face = "bold", hjust = 0.5))

ggplot(songs_2010s, aes(y = valence, x = acousticness, color = kclust_2010_scale)) +
    geom_point() +
    geom_smooth(se = FALSE) +
    scale_color_brewer(palette = "Set3") +
    labs(title = "2010s") +
    theme_classic()+
    theme(plot.background = element_rect(fill = "black", color = "black"), panel.background = element_rect(color = "black", fill = "black"), text = element_text(color = "white"), axis.text=element_text(color="white"), axis.line =element_line(color = "white"), legend.position = "none", plot.title = element_text(face = "bold", hjust = 0.5))
```

## Example from Popular Artists

```{r, message=F}
songs_artists <- songs %>%
  filter(performer %in% c("Beyonce", "Coldplay","Taylor Swift")) %>%
  select(year, song, performer, loudness, acousticness, valence)

songs_artists %>%
  ggplot(aes(x = year, y = loudness, color = performer)) +
  geom_point(size = 2, alpha = 0.7) + 
  geom_smooth(se = FALSE) +
  geom_vline(xintercept = 2010, color = "white", linetype = "dashed") +
  geom_hline(yintercept = -10, color = "white", linetype = "dotdash") +
  scale_color_manual(values = c("pink", "#FDE725", "#1F968B")) +
  theme_classic() +
  theme(plot.background = element_rect(fill = "black", color = "black"), panel.background = element_rect(color = "black", fill = "black"), legend.background = element_rect(color = "black", fill = "black"), text = element_text(color = "white"), axis.text=element_text(color="white"), axis.line =element_line(color = "white"), plot.title = element_text(size = 14, face = "bold", hjust = 0.5))

songs_artists %>%
  ggplot(aes(x = year, y = acousticness, color = performer)) +
  geom_point(size = 2, alpha = 0.7) + 
  geom_vline(xintercept = 2010, color = "white", linetype = "dashed") +
  geom_hline(yintercept = 0.5, color = "white", linetype = "dotdash") +
  scale_color_manual(values = c("pink", "#FDE725", "#1F968B")) +
  theme_classic() +
  theme(plot.background = element_rect(fill = "black", color = "black"), panel.background = element_rect(color = "black", fill = "black"), legend.background = element_rect(color = "black", fill = "black"), text = element_text(color = "white"), axis.text=element_text(color="white"), axis.line =element_line(color = "white"), plot.title = element_text(size = 14, face = "bold", hjust = 0.5))

songs_artists %>%
  ggplot(aes(x = year, y = valence, color = performer)) +
  geom_point(size = 2, alpha = 0.7) + 
  geom_smooth(se = FALSE) +
  geom_vline(xintercept = 2010, color = "white", linetype = "dashed") +
  geom_hline(yintercept = 0.5, color = "white", linetype = "dotdash") +
  scale_color_manual(values = c("pink", "#FDE725", "#1F968B")) +
  theme_classic() +
  theme(plot.background = element_rect(fill = "black", color = "black"), panel.background = element_rect(color = "black", fill = "black"), legend.background = element_rect(color = "black", fill = "black"), text = element_text(color = "white"), axis.text=element_text(color="white"), axis.line =element_line(color = "white"), plot.title = element_text(size = 14, face = "bold", hjust = 0.5))
```

